---
title: "lab_02"
author: "derek willis"
date: "2024-08-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

For this lab we'll be working with a slice of Maryland voter registration data: records of people who have registration dates this year through Aug. 7. Let's load it.

```{r}
md_voters_2024 <- read_rds("data/lab_02.rds")
```

## Examine Data

Take a look at the data, getting familiar with what it contains. Most of the columns should be self-explanatory, but there are two date columns: STATE_REGISTRATION_DATE and COUNTY_REGISTRATION_DATE. The former represents the date the person registered as a Maryland voter; the latter is when they registered in their current county. They may be the same, but if a voter moves from one county to another they will differ. This is important because if the two dates are the same, that's a pretty good indication that this is a *new* voter, someone who has no history of registering in Maryland. You'll need to keep that definition in mind.

## Answer questions

**Q1** How many voters appear to be *new* registrants since July 21, when President Joe Biden announced he would leave the race, based on the information above? Create a dataframe to hold the only those voters. Which counties have seen the most? Take a look at county population figures here and compare the two in your answer: https://worldpopulationreview.com/us-counties/maryland.

**A1** I created a way to see if voters are new registrants and it came back with over 17,000 people. Most of the counties in the map saw population growth but not by a large percent.

```{r}
md_voters_2024 |>
  filter(COUNTY_REGISTRATION_DATE > ymd("2024-7-21")) |>
  group_by(COUNTY) |>
    summarise(
      total = n()
    ) |>
  arrange(desc(total))
```

**Q2** A better question is whether the new voters you found in Q1 are an outlier compared to other weeks this year. Using the original dataframe with 2024 registrants, use `mutate()` and lubridate to produce week numbers from the COUNTY_REGISTRATION_DATE (Lubridate has a function that turns a date into a week number. Use that to answer this question). Save that to the original dataframe. Then calculate how many voters are in each week. Describe the distribution of voters groups in a sentence - are the weeks from July 21-August 7 different from the rest, and if so, how?

**A2** The second highest week is 7/21 with 7,126 people and the week after (7/28) is the third highest. The outlier is 6/9 with 7,531 people.

```{r}
md_voters_2024 <- md_voters_2024 |>
  mutate(week = week(COUNTY_REGISTRATION_DATE),
         week_start_date = floor_date(COUNTY_REGISTRATION_DATE, "week")
         )
md_voters_2024 |>
  group_by(week, week_start_date) |>
  summarise(
    count = n()
  ) |>
  arrange(desc(count))
```

**Q3** Using lubridate, show which month saw the largest number of registrations. Does the answer help shed more light on Q2?

**A3** July had the most number of registrants by a lot. This data is much easier to read than the data in question 2.

```{r}
md_voters_2024 |>
  mutate(month = month(COUNTY_REGISTRATION_DATE)) |>
  group_by(month) |>
  summarise(
    count = n()
  )
```

**Q4** Maryland has, officially, five regions: https://www.visitmaryland.org/article/maryland-regions. Modify the original dataframe again to add a `region` column based on the `COUNTY` value using case_when. Are you going to have to do this for every Maryland county plus Baltimore City? Yes, you are. Which regions and parties account for the most new voters?

**A4** I'm not sure I did this one right.

```{r}
md_voters_2024 |>
  mutate(region = case_when(
    COUNTY == "MONTGOMERY" ~ "REGION 1",
    COUNTY == "PRINCE GEORGE'S" ~ "REGION 12",
    COUNTY == "BALTIMORE" ~ "REGION 3",
    COUNTY == "BALTIMORE CITY" ~ "REGION 4",
    COUNTY == "ANNE ARUNDEL" ~ "REGION 5",
    COUNTY == "HOWARD" ~ "REGION 6",
    COUNTY == "FREDERICK" ~ "REGION 7",
    COUNTY == "HARFORD" ~ "REGION 8",
    COUNTY == "CARROLL" ~ "REGION 9",
    COUNTY == "CHARLES" ~ "REGION 10",
  ))
```

**Q5** What's an additional question you could ask of this data?

**A5** I would like to see if people provided a reasoning for when they registered to vote.
