---
title: "lab_03"
author: "Derek Willis"
date: "2024-08-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse
library(tidyverse)

# Load janitor
library(janitor)
```

## Load data

```{r}
# Load MD slice of FEC House candidate contribution data
md_contribs <- read_rds("data/lab_03.rds")
```

## Fix header issues

You'll need to fix one issue with the column names using janitor. Write a code block to do that, making sure that you save the results to the same dataframe.

```{r}
cleaned_md_contribs <- md_contribs |>
  clean_names()

cleaned_md_contribs
```


## Answer questions

Q1: Which seven cities in Maryland had the largest *number* of contributions from donors? How many did each have? To answer this question, you'll have to clean the messy "contributor_city" field.

Note: To answer this question, you'll need to see every variation of the "contributor_city" for the top seven cities, which you should identify through some initial analysis. Then find variations of them and change those variations into the most common version.

A1: Bethesda had the largest number of contributions with 1,213. The next six were Chevy Chase (705), Baltimore	(691), Potomac (517), Annapolis	(493), Silver Spring (408), and Rockville	(358). Many of them had names mispelled or duplicates that had to be cleaned.

```{r}
cleaned_md_contribs <- md_contribs |>
  clean_names() |>
  mutate(contributor_city = as.character(contributor_city)) |>
  mutate(contributor_city = str_to_title(contributor_city)) |>
  mutate(contributor_city = case_when(
    contributor_city == "Aaaaaa Oxon" ~ "Oxon Hill",
    contributor_city == "Balrimore" ~ "Baltimore",
    contributor_city == "4615 North Park Ave Apt. 1410" ~ "Chevy Chase",
    contributor_city == "Anapolis" ~ "Annapolis",
    contributor_city == "Annapolis Junction" ~ "Annapolis",
    contributor_city == "Silver Springs" ~ "Silver Spring",
    contributor_city == "North Potomac" ~ "Potomac",
    contributor_city == "Luthvle Timon" ~ "Timonium",
    contributor_city == "Lutherville Timoni" ~ "Timonium",
    contributor_city == "Lutherville Timonium" ~ "Timonium",
    contributor_city == "North Bethesda" ~ "Bethesda",
    contributor_city == "N Bethesda" ~ "Bethesda",
    TRUE ~ contributor_city
  )) |>
  group_by(contributor_city) |>
  summarise(count = n()) |> 
  arrange(desc(count)) |>
  distinct()

cleaned_md_contribs
```

Q2: Find the 5-digit ZIP code in Maryland that had the highest total amount of contributions (by dollar value). You will need to use mutate to ensure that you have a zip code column with the first five characters from the existing zip code column.

Where is that zip code and what kind of area is it, demographically? Which committee is the largest recipient of money from that zip code? How many of the top five recipients are Maryland candidates?

A2: The zip code is 20815. That zip code is Chevy Chase which is an area with an average income of over 200k a year and is predominantly white. It looks like Jamie Raskin is the top recipient who is from Maryland. The next four, April Mcclain Delaney (MD), Dan Helmer (VA), Ashley Ehasz (PA), and Eugene Vindman (PA).

```{r}
cleaned_md_contribs <- md_contribs |>
  clean_names() |>
 
mutate(contributor_zip = str_sub(contributor_zip, 1, 5)) |>
  mutate(contributor_zip = str_pad(contributor_zip, width = 5, pad = "0")) |>
  
  group_by(contributor_zip, committee_name) |>
  summarise(count = n()) |>
  filter(contributor_zip == "20815") |>
  arrange(desc(count))
  
cleaned_md_contribs
```

Q3: Let's look at top individual donors. Write code to generate a list of top donors using name and city, adding up the total amount of contributions. What does the original data say about the top two donors - who are they and what do they do? Do you see anything in the list of top 10 donors that would make you uncertain about the accuracy of these results?

A3: The top donor is Daniel Kracov from Rockville and the other is Margaret Allen from Baltimore. Kracov is the chair of a law firm and Allen owns a financial service. There are two people in the top ten with the same last name. Also, none of them are from Chevy Chase.

```{r}
cleaned_md_contribs <- md_contribs |>
  clean_names() |>
  
group_by(contributor_name, contributor_city) |>
 summarise(count = n()) |>
  arrange(desc(count))
  
cleaned_md_contribs
```
