---
title: "lab_12"
author: "derek willis"
date: "11/25/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

* Our usual libraries for working with data, plus rvest and possibly readxl for reading Excel files.

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.

```{r}
# Turn off scientific notation
options(scipen=999)

library(readxl)
library(tidyverse)
library(janitor)
library(rvest)
library(lubridate)
```

## Getting the Data

We're continuing our contrasting of the 2022 and 2018 elections, but this time we're switching to a different state: Utah. Utah had an unusual Senate election in 2022: Democrats did not nominate a candidate to run against incumbent Republican Mike Lee and instead endorsed an independent, Evan McMullin, in the race. Broadly, we want to dig into the differences between the 2022 Senate results and the 2018 Senate race, when Republican Mitt Romney defeated Democrat Jenny Wilson.

Let's start with the 2022 results. You'll need to locate and scrape the 2022 county-level results for the U.S. Senate race from the Utah Elections site (https://electionresults.utah.gov/elections/). Utah has 29 counties, so if you don't have at least 29 rows in your dataframe, you've scraped the wrong page.

**Task** Scrape the county-level results for the U.S. Senate race in 2022, which can be found at https://electionresults.utah.gov/elections/, into a dataframe. You will need to make sure that you have clean headers, which may involve renaming them, and you will need to remove the total row. Make sure that the columns containing votes are numeric.

```{r}
results_2022 <- read_csv("data/2022_utah_senate.csv")

colnames(results_2022)[1] <- "county"

clean_results_2022 <- results_2022 |>
  clean_names() |>
  mutate(mcmullin = as.numeric(mcmullin),
         hansen = as.numeric(hansen),
         lee = as.numeric(lee),
         williams = as.numeric(williams),
         county = str_to_lower(str_trim(county)),
         county = str_replace(county, " county$", ""))

clean_results_2022
```

Next, we'll get the 2018 results. Unlike the 2022 results, these are stored in an Excel file located here: https://elections.utah.gov/Media/Default/2018%20Election/2018%20General%20Election%20Canvass.xlsx. You can choose how to get the U.S. Senate results - whether you download and save the file in this lab_13 folder (we did this in pre_lab_12) or whether you download it manually and create a CSV file from the Senate results sheet in this directory. Like the 2022 results, you'll want to make a dataframe with clean headers, no total row and numeric columns for the votes.

**Task** Get the county-level results for the U.S. Senate race in 2018, which can be found at https://elections.utah.gov/Media/Default/2018%20Election/2018%20General%20Election%20Canvass.xlsx, and turn them into a dataframe. You will need to make sure that you have clean headers, which may involve renaming them, and you will need to remove the total row. Make sure that the columns containing votes are numeric.

```{r}
results_2018 <- read_csv("data/senate_results_2018.csv")

clean_results_2018 <- results_2018 |>
  clean_names() |>
  mutate(aalders = as.numeric(aalders),
         bowden = as.numeric(bowden),
         mccandless = as.numeric(mccandless),
         wilson = as.numeric(wilson),
         romney = as.numeric(romney),
         county = str_to_lower(str_trim(county)),
         county = str_replace(county, " county$", ""))

clean_results_2018
```

Finally, join the two dataframes together:

**Task** Join the 2022 and 2018 results into a single dataframe.

```{r}
clean_results_2018 <- clean_results_2018 |>
  mutate(county = str_squish(tolower(county)))

clean_results_2022 <- clean_results_2022 |>
  mutate(county = str_squish(tolower(county)))

combined_results <- clean_results_2018 |>
  left_join(clean_results_2022, by = "county")

combined_results
```

## Questions

**Q1.** Calculate the difference in votes between Evan McMullin and Jenny Wilson and save it your combined dataframe. Write up some sentences that could go in a story describing where McMullin outperformed Wilson and where he didn't. Mention the margins and describe the counties you mention (location, population or another factor, such as 2020 presidential vote).

**A1.** McMullin's most significant victories came in Utah County, Salt Lake County, and Davis County. In Utah County, McMullin's margin of victory was largest, with a difference of 43,382 votes. This is because it is one of the most conservative counties in Utah, partly due to its religious demographic, including members of The Church of Jesus Christ of Latter-day Saints, who tend to lean Republican.

```{r}
combined_results <- combined_results |>
  mutate(mcmullin_wilson_diff = mcmullin - wilson)

combined_results
```

**Q2** Calculate the difference in votes between Mike Lee and Mitt Romney and save it your combined dataframe. Write up some sentences that could go in a story describing where Romney outperformed Lee and where he didn't. Mention the margins and describe the counties you mention (location, population or another factor).

**A2** Mitt Romney outperformed Mike Lee most notably in Salt Lake County, the most populous county in the state, where he secured a margin of 62,788 votes. Romney’s performance was bolstered by strong support in the suburban and affluent areas, including Draper, South Jordan, and Herriman, where voters are more likely to align with Romney’s moderate conservative stance.

Meanwhile, Mike Lee found his strongest support in Washington County, where he outperformed Romney by 6,818 votes. Washington County, located in the southern part of Utah, is known for its more conservative, evangelical population.

Lee’s performance in Washington County reflects his strong appeal among rural and more right-wing voters, who often favor his conservative policies while Romney’s success in a more populated area highlights his appeal to centrist voters.

```{r}
combined_results <- combined_results |>
  mutate(lee_romney_diff = lee - romney)

combined_results
```

**Q3** Sen. Ben Cardin, D-Maryland, has posted hundreds of press releases at https://www.cardin.senate.gov/?post_type=press-releases. It would be great to have all of them in a dataframe that has the following columns: date, title and url.

To do this, you will need to scrape the page's html and save that to a variable, and separately use that variable to then extract the dates, titles and urls into separate dataframes using html_elements(). The function `html_text()` pulls out the contents of a tag, but for urls we want the HTML attribute. Rvest gives you a way to extract the URL from a link; google to find out what it is.

At the end, you'll have three dataframes that you want to combine into a single dataframe. When we want to combine the rows of identical dataframes, we used `bind_rows()`. There's an alternative you can use to do the same for combining columns, and it works the same way.

When you're done, rename the columns so they make sense, then make sure the date column is an actual date.

Finally, tell me what questions you could ask of this data. Be creative.

**A3** There are two questions that I immediately thought of to ask this data that could potentially lead to an interesting story. First, what are the most common topics addressed in Ben Cardin’s press releases? By looking for key words or phrases in the titles and content of the press releases, I could classify them into categories such as healthcare, economy, national security, or local issues. This could lead to a story about trends in his legislative priorities and areas of focus. Second, how does the frequency of press releases vary over time? While it will probably show that they are increased during election time, it could also show a correlation with world events, periods of intense legislative activity, etc. This could reveal how Cardin and his team handle serious world events and how he informs his supporters/Maryland.

```{r}
cardin_releases <- read_html("https://www.cardin.senate.gov/?post_type=press-releases")

dates <- cardin_releases |>  
  html_elements("h5") |> 
  html_text() |> 
  as_tibble()

titles <- cardin_releases |> 
  html_elements("h3") |> 
  html_text(trim = TRUE) |> 
  as_tibble()

urls <- cardin_releases |> 
  html_elements("h3 a") |> 
  html_attr("href") |> 
  as_tibble()

press_releases <- dates |>
  bind_cols(titles, urls) |>
  clean_names() |>
  rename(date = value_1,
         title = value_2,
         url = value_3)

press_releases
```
